CODECHECK: An open-science initiative for the independent execution of computations underlying research articles

### 2021-01-25 codecheck-paper.tex
fix last few TODOs

The first theme was the initial set of papers used to explore the concept
of CODECHECK. 
The idea was to take very well known historical articles from a 
specific domain, in this case the work area of SJE, possibly 
as as a conversation starter for potential collaborating journals.

The second theme is rather opportunistic. As the COVID-19 pandemic swept
over the whole Earth, we hoped to contribute a small part to science's answer
to the unknown virus. The checks were solicited through community interaction
%Several COVID-19 models have been done during the time of peer review of these papers,
and under our initiative rather than requested from journal, but some certificates
have since then been acknowledged in reviewed papers
\cite{Davies2020-vj,kucharski_effectiveness_2020}.

@@

Out of these areas, we would like to briefly highlight three examples here.
\textbf{TODO SJE: tidy this paragraph.}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Gigascience -- our first paper.  Only visualisation, not machine
  learning (which would have taken several days of compute time).
\item
  report 9 -- covid model of UK lockdown model -- unlike many claims
  in the popular media at the time, the model was reproducible
  [NOORDEN news piece]. 
\item
  Manuel's certificate -- approached us as he requested a certificate
  as paper was entering peer review.
\end{enumerate}

@@

The journal \emph{ACM Transactions on Mathematical Software (TOMS)} 
established a "Replicated Computational Results" (RCR) review process
\cite{heroux_editorial_2015} (using "replicable" with the same meaning 
as CODECHECK uses "reproducible").
A search on Crossref\footnote{\url{https://search.crossref.org/?q=\%22Replicated+Computations+Results+\%28RCR\%29+Report\%22&from_ui=yes&page=2}; search executed
on 2020-12-10.} shows 15 RCR Reports to be published and the 
% Number of found TOMAS RCR reports:
% 8: https://scholar.google.de/scholar?start=10&q=Replicated+Computations+Results+%22(RCR)+Report%22&hl=de&as_sdt=0,5
% 15: https://search.crossref.org/?q=Replicated+Computations+Results+%22%28RCR%29+Report%22&from_ui=yes&publication=ACM+Transactions+on+Mathematical+Software
% 15: https://academic.microsoft.com/search?q=Replicated%20Computations%20Results%20%22(RCR)%20Report%22&f=&orderBy=0&skip=10&take=10
workflow to be 
extended to the ACM journal \emph{Transactions on Modeling and Computer 
Simulation (TOMACS)}\footnote{\url{https://dl.acm.org/journal/tomacs/author-guidelines}}.

The CODECHECK principles were independently developed as we only later 
learned about TOMS RCR, which we judge as fulfilling CODECHECK principles
1 to 4.
This work provides an extensive discussion of the possible variants, 
while TOMS RCR realises one particular workflow for reproducibility 
reviews for one particular journal. The TOMS editorial
\cite{heroux_editorial_2015} shares similar concerns about
selection of reviewers as above. Unlike existing CODECHECK certificates, 
the RCR reports undergo editorial review, but this could very well become
part of CODECHECK processes at journals.
The publication of the RCR report is seen as a reward for the efforts of
the reproducing person, while the possible conflict of interest of this 
motivation is acknowledged. TOMS also recognises reviewer activity in a 
partnership with Publons\footnote{
\url{https://authors.acm.org/author-services/publons}},
though it is unclear if the author of an RCR report receives both the report's
authorship and the reviewer recognition. In any case, this is different
from CODECHECK, which specifically aims at recognition as reviewers.
In our view, the reviewer role removes the possible conflict of interest while
keeping the public acknowledgement.
Specific to the field of Mathematics, the RCR is also expected to apply a
review of the software itself if the system it runs on cannot be evaluated by
an independent party.
The TOMS RCR creators concur with the importance of communication, expect
collaboration between author and RCR reviewers,
share the considerations around reviewer selection, and also put trust
in reviewer judgement over numerical bit-wise perfection.
A key difference is that authors opt-in with an \emph{RCR Review Request} and
the RCR reports, which are very similar to CODECHECK certificates, are
published in the TOMS journal next to the actual papers.

@@

\textbf{CODECHECK requirements are not demanding enough}:
CODECHECK by design does not require authors to provide (and sustain) an
eternally functional workflow and neither suggests a specific software stack
or practical approaches, such as research compendia
\footnote{\url{https://research-compendium.science/}},
even if they are very valuable.
Creating something that anyone can reproduce has been called a 
fool's errand
\footnote{\url{https://twitter.com/DougBlank/status/1135904909663068165?s=09}}
and we tend to agree.
However, the package of data, code, and documentation
collaboratively created by authors and codecheckers is a snapshot of a 
working analysis that greatly increases the likelihood of a successful 
reproduction and possibility to extensions by third parties, with access
to suitable resources and matching skill set, in the future.
Concrete implementations of CODECHECK workflows, especially for specific
disciplines, may reify much more helpful guidelines for authors how to
create reproducibility packages.
The author-friendly ``low bar'' should not stay low forever, but cultural
change takes time and the encouragement and guidance that CODECHECK,
as part of the widely accepted peer review concept, can provide may
eventually allow to raise the bar as high as possible, e.g.,~with
executable research compendia \cite{nust_opening_2017}
or continuous analysis \cite{beaulieu-jones_reproducibility_2017-1}.
However, considering that missing artefacts and lack of documentation
have repeatedly been identified as main barriers to reproducibility
(e.g.,~\cite{stagge_assessing_2019,nust_improving_2020}),
we would not underestimate the power of a simple check.

@@

\begin{itemize}
\item TODO SJE: modelDB principle: minimum of one figure should be reproducibe.
  Example certificate \texttt{2020-016} (I think) which had only a small subset
  of the possible figures. Still useful and worth sharing.
\end{itemize}

>>>>

The first theme was the initial set of papers used to explore the concept
of CODECHECK. 
The idea was to take very well known historical articles from a 
specific domain, in this case the work area of SJE, possibly 
as as a conversation starter for potential collaborating journals.
This intention worked out with the first CODECHECK (certificate number
\texttt{2020-001}) before publication of an article for the journal
\emph{GigaScience}, which successfully reproduced the visualisations
though not the full machine learning process.

The second theme is rather opportunistic. As the COVID-19 pandemic swept
over the whole Earth, we hoped to contribute a small part to science's answer
to the unknown virus. The checks were solicited through community interaction
%Several COVID-19 models have been done during the time of peer review of these papers,
and under our initiative rather than requested from journal, but some certificates
have since then been acknowledged in reviewed papers
\cite{Davies2020-vj,kucharski_effectiveness_2020}. Unlike many claims in
popular media at the time,
% reference missing here TODO SJE, you had a note about some Noorden news piece
we found the results of Report 9, a model of the UK lockdown procedures,
to be reproducible (certificate number \texttt{2020-10}).

@@

@@

The journal \emph{ACM Transactions on Mathematical Software (TOMS)} 
established a "Replicated Computational Results" (RCR) review process
\cite{heroux_editorial_2015}, using "replicable" with the meaning 
as this work calls "reproducible".
A search on Crossref\footnote{\url{https://search.crossref.org/?q=\%22Replicated+Computations+Results+\%28RCR\%29+Report\%22&from_ui=yes&page=2}; search executed
on 2020-12-10.} shows 15 RCR Reports to be published and the 
% Number of found TOMAS RCR reports:
% 8: https://scholar.google.de/scholar?start=10&q=Replicated+Computations+Results+%22(RCR)+Report%22&hl=de&as_sdt=0,5
% 15: https://search.crossref.org/?q=Replicated+Computations+Results+%22%28RCR%29+Report%22&from_ui=yes&publication=ACM+Transactions+on+Mathematical+Software
% 15: https://academic.microsoft.com/search?q=Replicated%20Computations%20Results%20%22(RCR)%20Report%22&f=&orderBy=0&skip=10&take=10
workflow to be 
extended to the ACM journal \emph{Transactions on Modeling and Computer 
Simulation (TOMACS)}\footnote{\url{https://dl.acm.org/journal/tomacs/author-guidelines}}.
The CODECHECK principles were independently developed as we only later 
learned about TOMS RCR, which we judge as fulfilling CODECHECK principles
1 to 4.
This work provides an extensive discussion of the possible variants, 
while TOMS RCR realises one particular workflow for reproducibility 
reviews for one particular journal. The TOMS editorial
\cite{heroux_editorial_2015} shares similar concerns about
selection of reviewers as above. Unlike existing CODECHECK certificates, 
the RCR reports undergo editorial review, but this could very well become
part of CODECHECK processes at journals.
The publication of the RCR report is seen as a reward for the efforts of
the reproducing person, while the possible conflict of interest of this 
motivation is acknowledged. TOMS also recognises reviewer activity in a 
partnership with Publons\footnote{
\url{https://authors.acm.org/author-services/publons}},
though it is unclear if the author of an RCR report receives both the report's
authorship and the reviewer recognition. In any case, this is different
from CODECHECK, which specifically aims at recognition as reviewers.
In our view, the reviewer role removes the possible conflict of interest while
keeping the public acknowledgement.
Specific to the field of Mathematics, the RCR is also expected to apply a
review of the software itself if the system it runs on cannot be evaluated by
an independent party.
The TOMS RCR creators concur with the importance of communication, expect
collaboration between author and RCR reviewers,
share the considerations around reviewer selection, and also put trust
in reviewer judgement over numerical bit-wise perfection.
A key difference is that authors opt-in with an \emph{RCR Review Request} and
the RCR reports, which are very similar to CODECHECK certificates, are
published in the TOMS journal next to the actual papers.

@@

\textbf{CODECHECK requirements are not demanding enough}:
CODECHECK by design does not require authors to provide (and sustain) an
eternally functional workflow and neither suggests a specific software stack
or practical approaches, such as research compendia
\footnote{\url{https://research-compendium.science/}},
even if they are very valuable.
Creating something that anyone can reproduce has been called a 
fool's errand
\footnote{\url{https://twitter.com/DougBlank/status/1135904909663068165?s=09}}
and we tend to agree.
However, the package of data, code, and documentation
collaboratively created by authors and codecheckers is a snapshot of a 
working analysis that greatly increases the likelihood of a successful 
reproduction and possibility to extensions by third parties, with access
to suitable resources and matching skill set, in the future.
Concrete implementations of CODECHECK workflows, especially for specific
disciplines, may reify much more helpful guidelines for authors how to
create reproducibility packages.
The author-friendly ``low bar'' should not stay low forever, but cultural
change takes time and the encouragement and guidance that CODECHECK,
as part of the widely accepted peer review concept, can provide may
eventually allow to raise the bar as high as possible, e.g.,~with
executable research compendia \cite{nust_opening_2017}
or continuous analysis \cite{beaulieu-jones_reproducibility_2017-1}.
However, considering that missing artefacts and lack of documentation
have repeatedly been identified as main barriers to reproducibility
(e.g.,~\cite{stagge_assessing_2019,nust_improving_2020}),
we would not underestimate the power of a simple check.
ModelDB curation policies ensure manually to reproduce one figure only 
\cite{mcdougal_reproducibility_2016}, but that has not limited the usefulness
nor success of the platform.

@@
